---
title: "ch1_notes"
author: "William Hall"
date: "June 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Vectors

### Exercises

1) What are the six types of atomic vector? How does a list differ from an atomic vector?

  * dbl, int, char, logical, complex, raw
  * list can contain different types of atomic vectors. list have hierarchy. 

2) What makes is.vector() and is.numeric() fundamentally different to is.list() and is.character()?

  * is.vector() and is.numeric will return true for multiple different types of vectors. is.list() and is.character() will return true for only one type of vector. 

3) Test your knowledge of vector coercion rules by predicting the output of the following uses of c():

  * c(1, FALSE) > numeric 
  * c("a", 1) > character
  * c(list(1), "a") > list
  * c(TRUE, 1L) > numeric

    
4) Why do you need to use unlist() to convert a list to an atomic vector? Why doesn’t as.vector() work?

  * lists are vectors, and therefore as.vector() will not unlist a list 

5) Why is 1 == "1" true? Why is -1 < FALSE true? Why is "one" < 2 false?

  * 1 == "1" > the 1 gets converted to a character and thus the two vectors are equal.
  * -1 < FALSE > FALSE gets converted to a zero, and therefore -1 < 0 is true. 
  * "one" < 2 > 2 will get converted to a character and the comparison is based on char length?
  
  
6) Why is the default missing value, NA, a logical vector? What’s special about logical vectors? (Hint: think about c(FALSE,NA_character_).)

  * NA is logical because logical is lowest in the coercion hierarchy and therefor it will not result in other elements of the vectors getting coerced to something else.  

##Attributes and Factors

###Exercises

1. An early draft used this code to illustrate structure():

```{r}
structure(1:5, comment = "my attribute")
```

But when you print that object you don’t see the comment attribute. Why? Is the attribute missing, or is there something else special about it? (Hint: try using help.)

  * comment is a special attribute that is ignored by the print() function. 

2. What happens to a factor when you modify its levels?

```{r}
f1 <- factor(letters)
levels(f1) <- rev(levels(f1))
```

  * Modifying the levels changes the order in which the factor is displayed. 

 3. What does this code do? How do f2 and f3 differ from f1?

```{r}
f2 <- rev(factor(letters))
f3 <- factor(letters, levels = rev(letters))

```

  * f2: levels remain unchanged but the order of the factor is reversed. i.e. the element in the vector have been reversed. 
  * f3: the levels are changed but the elements in the vector remain unchanged. 

##Matrices and Arrays

###Notes

  * Arrays and matrices can be created with array() and matrix()
  * You also create a matrix with dim()

```{r}
x <- c(2,3,4,5)
dim(x) <- c(2,2)
x
```

   * Arrays can be n-dimmensional; matricies are two dimmensional. 
  * cbind() and rbind() are extensions of c()
  * t() lets you transpose a matrix. 
  
###Exercises

1. What does dim() return when applied to a vector?

```{r}
dim(x)
```

dim() return the dimmensions of a vector. i.e. row and column lengths. 

2. If is.matrix(x) is TRUE, what will is.array(x) return?

```{r}
is.matrix(x)
is.array(x)
```

3. How would you describe the following three objects? What makes them different to 1:5?

```{r}
x1 <- array(1:5, c(1, 1, 5)) %>% str()
x2 <- array(1:5, c(1, 5, 1)) %>% str()
x3 <- array(1:5, c(5, 1, 1)) %>% str()
```

They are n-dimmensional arrays. 

##Data frames

  * Data frames have the type list because they are lists with the same length.
  

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE)

typeof(df)
```

  * You can have a column that contains lists; this is because a data frame is of type list. 

```{r}
y <- list(c(1,2,3), c(5:9), c(10:12))
df$y <- y
```

  * However, when you pass a list to data.frame() the list elements will be treated as columns and this will cause problems if the lists are different lengths. 
  
  
```{r}
data.frame(x = 1:3, y = list(1:3, 2:3, 9:13))
```

* You can use I() to overcome this problem and make the list be treated as one element.

```{r}
data.frame(x = 1:3, y = I(list(1:3, 2:3, 9:13)))
```

##Exercises

1. 


